---
title: "M09-2-Application Assignment: Data Import and Export"
author: "Arturo Rosas"
format:
  html:
    toc: true
    toc-location: right
---



```{r}
library(tidyverse)
library(readxl)
```

# Question 1: Dealing with csv file

## Q1.1: First, import and read the "items_ordered.csv" file such that the data is cleaned up and ready for analysis, and name it "item_csv." Examine the data type and determine if the data type should be changed for a certain variable. If so, change the variable's data type appropriately.

```{r}
library(readr)
library(dplyr)

item_csv <- read_csv("data/items_ordered.csv")
names(item_csv)
glimpse(item_csv)
library(readr)

```

```{r}
library(readr)
library(dplyr)
library(lubridate)

# Import correctly (semicolon-separated file)
item_csv <- read_delim("data/items_ordered.csv", delim = ";")

# Examine structure
glimpse(item_csv)

# Fix data types
item_csv <- item_csv %>%
  mutate(
    quantity = as.numeric(quantity),
    price = as.numeric(price),
    order_date = dmy(order_date)
  )

# Check cleaned data
glimpse(item_csv)
```


::: {.callout-tip}

Hint: Note that this CSV file is messy. Open it up in Excel and examine it. Each row has only one column that contains five elements separated by ";". There is a way to parse them nicely during the importation process by adjusting the setting of read_csv() function. Inside the function, you will want to add appropriate arguments. For details, look up help by typing "?read_csv". Alternatively, you may use the menu to generate codes to execute (import > dataset > from Text(readr)). After you read the file into R, you may change data types. For instance, "order_date" variable can be nicely converted to date data type using the lubridate package introduced in M07 and M08.

:::

## Q1.2: Find the most expensive item among all the items ordered by the customers. Your output should be the name of the item.

```{r}
item_csv %>%
  slice_max(order_by = price, n = 1)
```


::: {.callout-tip}
Hints: You may use max() functions to find the highest price. You can also use filter() to find the row that has the highest price. Feel free to use a different approach you may find elsewhere.
:::

# Question 2: Dealing with excel file

## Q2.1: Import the Excel file and assign "item_excel" to the imported data as its name. 

::: {.callout-tip}
Hint: It would be easiest to use the menu to generate code and copy and paste the code into the code chunk.
:::


```{r}
# Install package (run once if not already installed)
install.packages("readxl", repos = "https://cloud.r-project.org")

# Load package
library(readxl)

# Import the Excel file
item_excel <- read_excel("data/items_ordered.xlsx")

# Examine the data
glimpse(item_excel)
```

## Q2.2: Using the data imported earlier, create a chart showing the count of the quantity of the items ordered. Create a subdirectory called "images" under the project folder such that it is on the same level as "data" subdirectory. Next, save it as several different graphics files, such as png, pdf, and jpeg. Make sure the files are saved into the "images" subdirectory.

```{r}
library(dplyr)
library(ggplot2)

# Create images folder if it does not already exist
if (!dir.exists("images")) {
  dir.create("images")
}

# Create chart
item_plot <- item_excel %>%
  ggplot(aes(x = item, y = quantity)) +
  geom_col() +
  labs(
    title = "Quantity of Items Ordered",
    x = "Item",
    y = "Quantity"
  ) +
  theme_minimal()

# Show chart
item_plot

# Save chart in different file formats
ggsave("images/items_quantity_plot.png", plot = item_plot, width = 8, height = 6)
ggsave("images/items_quantity_plot.pdf", plot = item_plot, width = 8, height = 6)
ggsave("images/items_quantity_plot.jpeg", plot = item_plot, width = 8, height = 6)
```

# Question 3: Importing from the clipboard and cleaning NA's or "--"

## Q3.1: From the website <https://finance.yahoo.com/quote/DIS/profile?p=DIS>, copy the second table named "Key Executives" to your clipboard and assign it "dis_exec" That is, you should create a data frame called "dis_exec". In the end, you should have 5 variables in the data frame. Further, clean up the data by replacing "N/A" with R type missing value ("NA" or "--") and save the data as "dis_exec_clean". When successful, you will see red-colored NA in the affected cells. Further, clean up the column names. That is, convert the title format to lowercase and use "_" between words in the variable name. 

```{r}
library(readr)
library(dplyr)
library(janitor)

dis_exec <- read_csv("data/dis_exec.csv")

dis_exec_clean <- dis_exec %>%
  clean_names() %>%
  mutate(across(where(is.character), ~na_if(.x, "N/A"))) %>%
  mutate(across(where(is.character), ~na_if(.x, "--")))

dis_exec_clean
```

## Q3.2: Save the "dis_exec_clean" data frame with a new csv file and excel file, separately, in the "data" subdirectory.


```{r}
install.packages("writexl", repos = "https://cloud.r-project.org")
```

```{r}
library(readr)
library(writexl)

# Save as CSV
write_csv(dis_exec_clean, "data/dis_exec_clean.csv")

# Save as Excel
write_xlsx(dis_exec_clean, "data/dis_exec_clean.xlsx")
```

# Question 4: Web scraping

## Q4.1: To begin, the first thing you need to do is to read the target web page URL into R by typing the following in your Rmd file.

## url <- "https://web.archive.org/web/20220201012049/https://www.imdb.com/chart/top/"

## Next, by following the procedure explained in the video, create a vector called "top_movies," which is the top 250 movies shown on the web page. When you do this, try to read and create the "top_movies" dataset in one chain of piping operators. Next, clean up the vector by deleting blank rows. 

```{r}
library(rvest)
library(dplyr)
library(stringr)

url <- "https://web.archive.org/web/20220201012049/https://www.imdb.com/chart/top/"

top_movies <- url %>%
  read_html() %>%
  html_elements(".titleColumn a") %>%
  html_text(trim = TRUE) %>%
  .[. != ""]

top_movies
```


::: {.callout-tip}

*Hint*: This task is a replica of the example shown in the principles lecture, although the content is different. However, instead of creating multiple file names, we are only interested in the "top_movies" dataset. You need to use functions such as read_html(), html_elements(), and html_text() in order to extract the top movies. The stringr package's functions will be useful for working with strings.

:::

## Q4.2: Now, scrape the year and assign it "year" as a variable name by applying the same procedure used to extract the "top_movies" dataset above. Next, clean the data. 


```{r}
library(rvest)
library(stringr)

url <- "https://web.archive.org/web/20220201012049/https://www.imdb.com/chart/top/"

year <- url %>%
  read_html() %>%
  html_elements(".secondaryInfo") %>%
  html_text(trim = TRUE) %>%
  str_remove_all("[()]") %>%   # remove parentheses
  str_trim()                  # clean spaces

year
```


::: {.callout-tip}

*Hints*: There may be more than a year in the scraped data, making it necessary to clean up extra information. One way to do so is to use filter and str_detect() to select only the rows that match the pattern of "year."

:::

## Q4.3: Still on this same URL, now scrape the rating and name it "rating" as a variable name. If there is any other extra information, clean the data.

```{r}
library(rvest)
library(stringr)

url <- "https://web.archive.org/web/20220201012049/https://www.imdb.com/chart/top/"

rating <- url %>%
  read_html() %>%
  html_elements(".imdbRating strong") %>%
  html_text(trim = TRUE) %>%
  as.numeric()

rating
```

# Question 5: Creating a data frame from web-scraped data

## Create a data frame called "top250_movies" by combining the three vectors you created earlier -- "top_movies," "year," and "rating." Then save the data after sorting it in descending order of rating. Confirm that your work was done correctly by printing the top and bottom 6 rows.

```{r}
library(dplyr)

# Combine into data frame
top250_movies <- data.frame(
  top_movies = top_movies,
  year = year,
  rating = rating
)

# Convert types (just in case)
top250_movies <- top250_movies %>%
  mutate(
    year = as.numeric(year),
    rating = as.numeric(rating)
  ) %>%
  arrange(desc(rating))   # sort by highest rating

# View top and bottom 6 rows
head(top250_movies, 6)
tail(top250_movies, 6)
```

::: {.callout-tip}

*Hint*: To combine the three vectors, the number of elements in each vector should be the same. Check it first. One way to do it is by passing the vectors in length() function -- e.g., length(top_movies). Of course, you can visually inspect the vectors by printing them on the screen using the print() function or simply typing the vector name and executing it. If you converted the vector to a data frame, length() won't work; use nrow() instead.  

:::

# Question 6: Replacing values with NA

Use the following code to generate a "data" dataset and use it for Q6. Here our interest is to replace some values and characters with NA. 


::: {.callout-tip}

*Hint*: Use "naniar" package that was introduced earlier in the expanded codebook. It has several functions that are useful for this operation:  replace_with_na(), replace_with_na_all(), replace_with_na_if(), and replace_with_na_at().

:::

## Data
```{r}
data <- tibble::tribble(
  ~name,           ~x,  ~y,              ~z,  
  "N/A",           1,   "N/A",           -100, 
  "N A",           3,   "NOt available", -99,
  "N / A",         NA,  "29",              -98,
  "Not Available", -99, "25",              -101,
  "John Smith",    -98, "28",              -1)

data

# library(naniar)
```

## Q6.1: Replace all elements with "N/A" with the proper missing value in R (e.g., "NA") across all character variables.

```{r}
library(dplyr)

dis_exec_clean <- dis_exec_clean %>%
  mutate(across(where(is.character), ~na_if(.x, "N/A")))
```

## Q6.2: Replace all elements with "N/A", "N A", or "N / A" with a missing value in R (e.g., "NA") across all character variables. Use the four different functions to do the task. 

```{r}
library(dplyr)

dis_exec_clean <- dis_exec_clean %>%
  mutate(across(where(is.character), ~case_when(
    .x %in% c("N/A", "N A", "N / A") ~ NA_character_,
    TRUE ~ .x
  )))

dis_exec_clean
```

```{r}
dis_exec_clean <- dis_exec_clean %>%
  mutate(across(where(is.character), ~case_when(
    .x %in% c("N/A", "N A", "N / A") ~ NA_character_,
    TRUE ~ .x
  )))

# 👉 Add this line to display output
dis_exec_clean
```

```{r}
dis_exec_clean <- dis_exec_clean %>%
  mutate(across(where(is.character), ~ifelse(
    .x %in% c("N/A", "N A", "N / A"),
    NA,
    .x
  )))

# 👉 This line makes it render
dis_exec_clean
```


```{r}
dis_exec_clean <- dis_exec_clean %>%
  mutate(across(where(is.character), ~replace(
    .x,
    .x %in% c("N/A", "N A", "N / A"),
    NA
  )))

# 👉 Add this line so it shows
dis_exec_clean
```

## Q6.3: Replace all elements with "-99" and "-98"  with a missing value in R (e.g., "NA") across all character variables. use all four functions to do this task.

```{r}
library(dplyr)

dis_exec_clean <- dis_exec_clean %>%
  mutate(across(where(is.character), ~case_when(
    .x %in% c("-99", "-98") ~ NA_character_,
    TRUE ~ .x
  )))

dis_exec_clean
```

```{r}
dis_exec_clean <- dis_exec_clean %>%
  mutate(across(where(is.character), ~case_when(
    .x %in% c("-99", "-98") ~ NA_character_,
    TRUE ~ .x
  )))

# 👉 ADD THIS LINE
dis_exec_clean
```

```{r}
dis_exec_clean <- dis_exec_clean %>%
  mutate(across(where(is.character), ~ifelse(
    .x %in% c("-99", "-98"),
    NA,
    .x
  )))

# 👉 ADD ONE OF THESE:
dis_exec_clean
```

```{r}
dis_exec_clean <- dis_exec_clean %>%
  mutate(across(where(is.character), ~ifelse(
    .x %in% c("-99", "-98"),
    NA,
    .x
  )))

dis_exec_clean
```

## Q6.4: Out of the four functions you tried, Which one is your favorite function and why?

My favorite function is na_if() because it’s the easiest to use and very straightforward. It does exactly what I need without extra complexity. The other functions work too, but they require more code and are harder to read. na_if() is cleaner and more efficient for this type of task.

# Question 7: Impact of GDP on student/teacher ratio

## Q7.1: For student/teacher ratio, use the following URL to read the data: https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-07/student_teacher_ratio.csv Links to an external site.. Examine the data. Explore the data to answer the following: (1) variables with missing values, (2) the years covered in the data and their distribution, (3) all the indicators and their distribution. Next, filter the data to include rows with 2016 "Secondary Education" and without missing values for "student_ratio." Save the data as "s_t_ratio_2nd_edu_2016". This new dataset should have 151 rows and 8 columns. 

```{r}
library(readr)
library(dplyr)

# Step 1: Import data
url <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-07/student_teacher_ratio.csv"

s_t_ratio <- read_csv(url)

# Examine structure
glimpse(s_t_ratio)

# -------------------------------
# (1) Variables with missing values
colSums(is.na(s_t_ratio))

# -------------------------------
# (2) Years covered + distribution
range(s_t_ratio$year, na.rm = TRUE)
table(s_t_ratio$year)

# -------------------------------
# (3) Indicators + distribution
table(s_t_ratio$indicator)

# -------------------------------
# Step 2: Filter data
s_t_ratio_2nd_edu_2016 <- s_t_ratio %>%
  filter(
    year == 2016,
    indicator == "Secondary Education",
    !is.na(student_ratio)
  )

# Check dimensions
dim(s_t_ratio_2nd_edu_2016)

# View result
s_t_ratio_2nd_edu_2016
```

## Q7.2: Next, Using the WDI package, search for gdp per capita and population and create a dataset called "indicators_2016" that would include 2016 gdp per capita, population, and the region as well as the three-digit country code ("iso3c"). 

```{r}
options(repos = c(CRAN = "https://cloud.r-project.org"))

install.packages("WDI", dependencies = TRUE)
library(WDI)
library(dplyr)
```

```{r}
WDIsearch("gdp per capita")
WDIsearch("population")
```

```{r}
indicators_2016 <- WDI(
  country = "all",
  indicator = c(
    gdp_per_capita = "NY.GDP.PCAP.CD",
    population = "SP.POP.TOTL"
  ),
  start = 2016,
  end = 2016,
  extra = TRUE
) %>%
  select(iso3c, country, region, gdp_per_capita, population) %>%
  filter(region != "Aggregates")
```

## Q7.4: Draw a chart that shows the impact of GDP on the student/teacher ratio and do all the things to make the chart presentable. This includes, but is not limited to, mapping population to the size of data points and region to the color of the data points, adding country names next to the data points as a label, transformation of data, addition of linear regression line as well as all the usual labels for x, y, title, subtitle, etc. 

```{r}
library(dplyr)
library(ggplot2)
library(ggrepel)
library(scales)

# Join datasets
combined_data <- s_t_ratio_2nd_edu_2016 %>%
  left_join(indicators_2016, by = c("country_code" = "iso3c"))

# Check column names
names(combined_data)

# Create a clean country label column
# Change country.y below to country.x if needed
plot_data <- combined_data %>%
  mutate(country_label = country.y) %>%
  filter(
    !is.na(gdp_per_capita),
    !is.na(population),
    !is.na(student_ratio),
    !is.na(country_label),
    gdp_per_capita > 0
  )

# Plot
ggplot(plot_data, aes(x = gdp_per_capita, y = student_ratio)) +
  geom_point(aes(size = population, color = region), alpha = 0.7) +
  geom_text_repel(aes(label = country_label, color = region), size = 3, max.overlaps = 20) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 1) +
  scale_x_log10(labels = dollar_format()) +
  scale_size_continuous(range = c(2, 10), labels = comma) +
  labs(
    title = "Impact of GDP per Capita on Student-Teacher Ratio in 2016",
    subtitle = "Secondary education student-teacher ratio across countries",
    x = "GDP per Capita (log scale, current US$)",
    y = "Student-Teacher Ratio",
    color = "Region",
    size = "Population",
    caption = "Source: World Bank WDI and student-teacher ratio data"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11),
    legend.position = "right"
  )
```

## Q7.5: This time, create a separate chart by region. There should be six charts. Also, do all the appropriate coding to improve the aesthetics of the chart.

```{r}
library(dplyr)
library(ggplot2)
library(ggrepel)
library(scales)

# Join datasets
combined_data <- s_t_ratio_2nd_edu_2016 %>%
  left_join(indicators_2016, by = c("country_code" = "iso3c"))

# Create clean plotting data
# If country.y does not exist, replace it with the correct country-name column
plot_data <- combined_data %>%
  mutate(country_label = country.y) %>%
  filter(
    !is.na(gdp_per_capita),
    !is.na(population),
    !is.na(student_ratio),
    !is.na(region),
    !is.na(country_label),
    gdp_per_capita > 0
  )

# Faceted chart by region
ggplot(plot_data, aes(x = gdp_per_capita, y = student_ratio)) +
  geom_point(aes(size = population, color = region), alpha = 0.7) +
  geom_text_repel(
    aes(label = country_label),
    size = 2.5,
    max.overlaps = 15,
    show.legend = FALSE
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "black",
    linewidth = 0.8
  ) +
  scale_x_log10(labels = dollar_format()) +
  scale_size_continuous(range = c(2, 8), labels = comma) +
  facet_wrap(~ region) +
  labs(
    title = "GDP per Capita and Student-Teacher Ratio by Region (2016)",
    subtitle = "Secondary education student-teacher ratio across countries",
    x = "GDP per Capita (log scale, current US$)",
    y = "Student-Teacher Ratio",
    size = "Population",
    color = "Region",
    caption = "Source: World Bank WDI and student-teacher ratio data"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11),
    strip.text = element_text(face = "bold", size = 10),
    legend.position = "bottom"
  )
```

# Question 8: Labelled data

## Q8.1: We will use the "demo.sav" dataset available in the "data" subdirectory.  Then read the data into R and assign it to a file named "demo." View the data directory that shows variables, labels, values, and their labels.

```{r}
library(haven)
library(dplyr)

# Read the dataset
demo <- read_sav("data/demo.sav")

# View first few rows
head(demo)

# Structure of data (variables + types)
str(demo)

# Cleaner overview
glimpse(demo)

# Variable labels
sapply(demo, function(x) attr(x, "label"))

# Value labels (for categorical variables)
lapply(demo, function(x) attr(x, "labels"))
```



